When the same sample is processed more than once (e.g. repeat runs of a tool
over different dates), you end up with several sets of raw outputs that share
**identical file names** but live under **different parent folders**. Here we
show: (a) how tidy outputs are named so that repeated runs avoid overwriting
each other, and (b) how to carry a run identifier *into the data* so the runs
are still distinguishable once the files have been merged into a parquet dataset
or database.

## Anatomy of a tidy filename

Every tidy file is named:

```
{output_dir}/{prefix}_{tool}_{parser}.{ext}
```

- `{output_dir}`: the directory to output the tidy files in a flat structure.
- `{prefix}`: the input file's basename with the table's schema `pattern`
  stripped off. For `sampleA.tool1.table1.tsv` the `table1` pattern
  (`\.tool1\.table1\.tsv$`) is removed, leaving `sampleA`.
- `{tool}_{parser}`: the tool name and the matched table (e.g. `tool1_table1`).
- `{ext}`: driven by `format`
  (`r nemo::nemo_out_formats() |> glue::glue_collapse(sep = ", ", last = " or ")`).

So `sampleA.tool1.table1.tsv` tidied in parquet format becomes
`sampleA_tool1_table1.parquet`.

## A repeated sample

Let us simulate three runs of the same sample by copying one tool's outputs into
separate folders. Each run holds the same files:

```{r}
#| label: build-runs

src <- system.file("extdata/tool1/latest", package = "nemo")
dir1 <- path(tempdir(), "naming-demo")
dir_runs <- path(dir1, "runs")
run_ids <- c("run1", "run2", "run3")

for (r in run_ids) {
  dest <- dir_create(path(dir_runs, r))
  file_copy(dir_ls(src, regexp = "table[12]\\.tsv$"), dest, overwrite = TRUE)
}

dir_tree(dir_runs)
```

### Mode A: one recursive tidy

Point a single `Tool1` at the *parent* of the run folders. `list_files()`
recurses into all three and, because the basenames collide, appends `_2` / `_3`
to disambiguate the prefixes:

```{r}
#| label: modeA-listfiles

tool <- Tool1$new(path = dir_runs)
tool$list_files() |>
  dplyr::select(bname, parser, prefix)
```

As you can see, the files share a basename but get prefixes `sampleA`,
`sampleA_2`, `sampleA_3`. Writing them out, nothing gets overwritten:

```{r}
#| label: modeA-write

dir_outA <- dir_create(path(dir1, "outA"))
tool$run(
  input_id = "input1",
  output_dir = dir_outA,
  format = "parquet",
  prefix_include = TRUE
)

dir_tree(dir_outA)
```

The `prefix_include` option also writes the prefix as an `input_prefix` column
in the output file itself, so the source run survives even after the three files
are read and stacked into one table:

```{r}
#| label: modeA-merge

dir_ls(dir_outA, regexp = "tool1_table1\\.parquet") |>
  purrr::map(arrow::read_parquet) |>
  purrr::list_rbind()
```

Mode A is the simplest way to dump every tidy run into one folder without
overwriting each other. Its limit: the tidy call shares a single `input_id`, so
the only per-run key is the filename-derived `input_prefix`.

### Mode B: per-run tidy

When you want a more meaningful run identifier than e.g. `sampleA_2`, you should
tidy each run separately and pass a distinct `input_id`. Because the output
basenames are now identical across runs, write each run to its own output
subfolder so they don't get overwritten:

```{r}
#| label: modeB-write

dir_outB <- dir_create(path(dir1, "outB"))
for (r in run_ids) {
  Tool1$new(path = path(dir_runs, r))$run(
    input_id = r,
    output_dir = path(dir_outB, r),
    format = "parquet",
    prefix_include = TRUE
  )
}

dir_tree(dir_outB)
```

Note the same file names in each subfolder. The folder separates them on disk,
and the `input_id` column keeps them apart once merged:

```{r}
#| label: modeB-merge

fs::dir_ls(dir_outB, recurse = TRUE, glob = "*tool1_table1.parquet") |>
  purrr::map(arrow::read_parquet) |>
  purrr::list_rbind()
```

For a globally-unique key with no coordination between runs, use
`output_id = "<your id>"` with e.g. an auto-generated ULID (done automatically
via the CLI's `--ulid` flag), which would generate an `output_id` column
alongside `input_id`.

## Rule of thumb

- File names stop repeat runs from overwriting each other on disk automatically
  via `_2`/`_3` (Mode A), or by writing to per-run folders (Mode B).
- Provenance columns (`input_prefix`, `input_id`, `output_id`) let you tell the
  runs apart after the files are merged into one parquet dataset or loaded into
  a database, where the filename is no longer visible.
- Use `prefix_include` when one recursive tidy is enough
- Use `input_id` / `output_id` when each run is better described by an
  explicitly specified name you can control.
