Repeat runs of the same sample produce identical file names in different
folders. Below: how tidy outputs avoid overwriting each other, and how to keep
runs distinguishable after merging.

## Anatomy of a tidy filename

```
{output_dir}/{prefix}_{tool}_{parser}.{ext}
```

```{r}
#| label: out-fmt
#| echo: false

fmts <- nemo::nemo_out_formats() |>
  glue::glue_collapse(sep = ", ", last = " or ")
```

- `{output_dir}`: flat output directory
- `{prefix}`: input basename minus the schema `pattern`
  (`sampleA.tool1.table1.tsv` => `sampleA`)
- `{tool}_{parser}`: e.g. `tool1_table1`
- `{ext}`: from `format` (`r fmts`)

`sampleA.tool1.table1.tsv` => `sampleA_tool1_table1.parquet`.

## A repeated sample

Three runs, same input files:

```{r}
#| label: build-runs

src <- system.file("extdata/tool1/latest", package = "nemo")
dir1 <- path(tempdir(), "naming-demo")
dir_runs <- path(dir1, "runs")
run_ids <- c("run1", "run2", "run3")

for (r in run_ids) {
  dest <- dir_create(path(dir_runs, r))
  file_copy(dir_ls(src, regexp = "table[12]\\.tsv$"), dest, overwrite = TRUE)
}

dir_tree(dir_runs)
```

### Mode A: one recursive tidy

Point `Tool1` at the parent folder. Colliding prefixes get `_2`/`_3`:

```{r}
#| label: modeA-listfiles

tool <- Tool1$new(path = dir_runs)
tool$list_files() |>
  dplyr::select(path, prefix, parser) |>
  dplyr::arrange(parser)
```

No overwrites:

```{r}
#| label: modeA-write

dir_outA <- dir_create(path(dir1, "outA"))
tool$run(
  input_id = "input1",
  output_dir = dir_outA,
  format = "parquet",
  prefix_include = TRUE
)

dir_tree(dir_outA)
```

`prefix_include = TRUE` adds an `input_prefix` column, so runs stay distinct
after stacking:

```{r}
#| label: modeA-merge

dir_ls(dir_outA, regexp = "tool1_table1\\.parquet") |>
  purrr::map(arrow::read_parquet) |>
  purrr::list_rbind()
```

Limit: one shared `input_id`; the only per-run key is `input_prefix`.

### Mode B: per-run tidy

Tidy each run separately with its own `input_id` and output subfolder:

```{r}
#| label: modeB-write

dir_outB <- dir_create(path(dir1, "outB"))
for (r in run_ids) {
  Tool1$new(path = path(dir_runs, r))$run(
    input_id = r,
    output_dir = path(dir_outB, r),
    format = "parquet",
    prefix_include = TRUE
  )
}

dir_tree(dir_outB)
```

Same file names per subfolder; `input_id` separates them once merged:

```{r}
#| label: modeB-merge

fs::dir_ls(dir_outB, recurse = TRUE, glob = "*tool1_table1.parquet") |>
  purrr::map(arrow::read_parquet) |>
  purrr::list_rbind()
```

For a globally-unique key, set `output_id`, e.g. to a [ULID][ulid] (CLI:
`--ulid`, via the [ulid][ulid-r] R package).

## Rule of thumb

- On disk: `_2`/`_3` suffixes (Mode A) or per-run folders (Mode B)
- After merging: `input_prefix`, `input_id`, `output_id` columns
- One recursive tidy is enough => `prefix_include`
- Want named runs => `input_id` / `output_id`

[ulid]: https://wiki.tcl-lang.org/page/ULID "What is ULID"
[ulid-r]: https://github.com/eddelbuettel/ulid "ULID in R"
