---
title: "NEWS"
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    echo: false
---

```{r}
#| label: pkgs
#| include: false

repo <- "https://github.com/tidywf/nemo"
child1 <- system.file(
  "doc-templates/vignettes/_news.qmd",
  package = "nemo"
)
```

```{r, child=child1}
#| label: child-news
```

## v0.1.1 (unreleased)

### New features

- `Tool` gains a `flat_tidy_names` field (default `FALSE`). When `TRUE`, a
  fanned-out (one-file-to-many-tables) sub-table is named `<tool>_<tidy_name>`
  instead of the concatenated `<tool>_<parser><tidy_name>`, giving flat output
  names (e.g. `dragenfastqc_posbasecontent`). The default preserves existing
  behaviour for every tool; `write()` errors if the dropped parser token would
  let two outputs collide to the same name.

### Breaking changes

- Removed the `csv-nohead-long` ftype and the associated `table5` fixture from
  the `tool1` example. Child packages needing DRAGEN-style long-metrics parsing
  should implement their own custom `parse_*`/`tidy_*` (see `tidydragen`).

## v0.1.0 (2026-07-28)

Major refactor.

### Breaking changes

**Unified schema format** (`r pr(48)`)

- `schema.yaml` replaces the separate `raw.yaml` + `tidy.yaml` pair per tool.
  Each column now carries `raw`, `tidy`, `type`, `description`, and `versions`
  in one place under a flat `tables:` map. Child packages must migrate their
  per-tool configs to this format.

**Config / Tool / Workflow API overhaul** (`r pr(64)`, `r pr(65)`, `r pr(66)`)

- `Config`: `get_raw_*` / `get_tidy_*` methods renamed (e.g.
  `get_raw_patterns()` → `get_patterns()`, `get_raw_schemas_all()` →
  `get_schemas_raw()`); `are_raw_schemas_valid()` removed (validation now in
  `initialize()`); `config`, `raw_schemas_all`, `tidy_schemas_all` fields moved
  to private; `pkg` added as a public field; `cloneable = FALSE`
- `Tool`: `$files`, `$tbls`, `$files_tbl` moved to private --- use
  `list_files()` / `get_tbls()` instead; schema delegate fields
  (`get_raw_schema`, `get_tidy_schema`) removed --- use
  `tool$config$get_schema_raw()` / `tool$config$get_schema_tidy()`;
  `nemofy(diro = ...)` → `run(output_dir = ...)`; column `group` →
  `prefix_suffix`; `post_process_files` hook renamed to `refine_files`, now
  applied to the base prefix *before* the disambiguation passes so subclasses
  can fold a semantic distinction (e.g. germline/somatic) into the prefix and
  avoid a spurious `_2`/`_3` suffix; `cloneable = FALSE`
- `Workflow`: `$tools`, `$files_tbl` moved to private --- use `get_tools()`;
  `nemofy(diro = ...)` → `run(output_dir = ...)`; `list_files(type = ...)` ---
  `type` arg removed; gains `metapkg` constructor argument; path existence
  validated at construction; `filter_files()` now validates names against known
  `tool_parser` values; `cloneable = FALSE`

**CLI renames**

- `--out_dir` → `--output_dir`; `--pfix_include` → `--prefix_include`; output
  column `input_pfix` → `input_prefix`; `--prefix_include` is now opt-in (no
  longer added by default)

**Other**

- `nemo_metadata()`: parameter `input_dir` => `input_dirs`; return value changed
  from a named list to a single-row tibble with list-columns
- `RPostgres` dropped as a hard dependency - callers now supply `dbdrv`
  explicitly
- `txt` ftypes renamed to `tsv` (`r pr(70)`, `r pr(71)`)

### New features

- New assertion module: `nemo_stop()`, `nemo_assert_scalar_chr()`,
  `nemo_assert_chr()`, `nemo_assert_not_null()`, `nemo_assert_out_fmt()`
  replacing `assertthat` calls
- Schema scaffold helpers: `config_prep_raw_schema()`, `config_prep_raw()`,
  `config_prep_multi()`, `config_prep_write()` --- bootstrap a `schema.yaml`
  from example raw files
- `nemo_gha_mermaid()`: generates a Mermaid flowchart of the full CI/CD pipeline
  (`r pr(52)`)
- `nemo_uml()`: generates a PlantUML SVG from R6 class names
- `nemo_schema_reactable()` / `nemo_schemavis_data()`: interactive schema
  explorer
- New file types: `tsv-nohead`, `tsv-keyvalue`, `csv`, `csv-nohead-long`
- `tidy`: `--output_id` / `--ulid` flags for tagging outputs with a run
  identifier; `--max` flag on `list` subcommand
- Metadata exported to `metadata.parquet` per run (`r pr(62)`, `r iss(61)`)
- `Tool` and `Workflow` now both support `write_metadata` boolean
- `extra_ftypes()` subclass hook for registering pkg-specific ftype parsers
  (`r pr(70)`, `r pr(71)`)
- Add [DVC](https://dvc.org/) support, mostly for child packages (`r pr(69)`)
- New `Output Naming` vignette, shared with child packages via
  `inst/doc-templates/vignettes/_output_naming.qmd` (`r pr(73)`)

### Performance

- File discovery \~20× faster on large directories --- benchmarked at 1.9s =>
  0.1 s on 43k files x 15 patterns

### Infrastructure

- CI/CD: `deploy.yaml` refactored from a 136-line monolith to reusable workflows
  via `tidywf/actions`; now also triggers on `dev` branch pushes
- `@claude` GHA restricted to repo owners, collaborators, and members
- `dependabot.yml` added for automated GHA dependency updates
- conda: aarch64 lock file added; dedicated `bump.yaml` env added
- Panache format vignettes (`r pr(72)`)

## v0.0.3 (2026-02-09)

- `nemo_write`: use NA outpath attr for db

## v0.0.2 (2026-02-04)

- Add key-value parser (`r pr(12)`, `r iss(11)`)
- Export config writer (`r pr(13)`)
- Add table description to schema getters, contribution docs, pkgdown fixes
  (`r pr(19)`)
- Add `nemo_id` + `nemo_pfix` in class writer (`r pr(21)`, `r iss(18)`)
  - changed odir to diro arg in nemofy
- list_files: `str_remove` offers more flexibility (`r pr(24)`, `r iss(23)`)
- Add metadata export functionality (`r pr(25)`, `r iss(17)`)
  - rename `nemo_id` + `nemo_pfix` to `input_id` + `input_pfix`, respectively
  - add `output_id`
- Bugfix: use `I()` for vroom literal data in `empty_tbl` (`r iss(27)`,
  `r pr(28)`)

## v0.0.1 (2025-09-07)

Initial release of nemo (`r pr(5)`)

- R pkg skeleton
- Documentation via pkgdown
- CLI
- Logger
- GitHub Actions CI/CD

Fixing issues `r iss(2)`, `r iss(3)` and `r iss(4)`
