---
title: "NEWS"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
    echo: false
---

```{r}
#| label: pkgs
#| message: false
#| warning: false
{
  use("htmltools", "a")
  use("glue", "glue")
}
```

```{r}
#| label: setup
repo <- "https://github.com/tidywf/nemo"
vdiff <- function(new, old) {
  stopifnot(utils::compareVersion(new, old) == 1)
  htmltools::a(
    href = glue::glue("{repo}/compare/v{old}...v{new}"),
    glue::glue("v{new} - v{old} diff")
  )
}

pr_or_issue <- function(n, type) {
  x <- c(iss = "issues", pr = "pull")
  stopifnot(is.numeric(n), n > 0, type %in% names(x))
  type2 <- x[type]
  htmltools::a(
    href = glue::glue("{repo}/{type2}/{n}"),
    glue::glue("{type}{n}")
  )
}

iss <- function(n) {
  pr_or_issue(n, "iss")
}

pr <- function(n) {
  pr_or_issue(n, "pr")
}
```

## v0.1.0 (2026-06-04)

### Breaking changes

- Unified schema format: `schema.yaml` replaces separate `raw.yaml` + `tidy.yaml`
  per tool. Each column now carries its raw name, tidy name, type, description,
  and applicable versions in one place
- CLI: `--out_dir` → `--output_dir`; `--pfix_include` → `--prefix_include`;
  output column `input_pfix` → `input_prefix`
- `Config`, `Tool`, and `Workflow` APIs overhauled — methods and fields renamed
  for consistency across the package

### New features

- `tidy`: add `--output_id` / `--ulid` for tagging outputs with a run identifier
  (`--ulid` auto-generates one)
- `tidy`: `--prefix_include` is now opt-in — `input_prefix` is no longer added
  to outputs by default
- New file types supported by `Tool`: `txt-nohead`, `csv-nohead-long`, `csv`
- Metadata exported to `metadata.parquet` per run (`r pr(62)`, `r iss(61)`)
- `nemo_gha_mermaid()`: generates a Mermaid flowchart of the full CI/CD pipeline
  (`r pr(52)`)

### Performance

- File discovery ~20× faster on large directories — benchmarked at 1.9 s → 0.1 s
  on 43k files × 15 patterns

## v0.0.3 (2026-02-09)

- `nemo_write`: use NA outpath attr for db

## v0.0.2 (2026-02-04)

- Add key-value parser (`r pr(12)`, `r iss(11)`)
- Export config writer (`r pr(13)`)
- Add table description to schema getters, contribution docs, pkgdown fixes (`r pr(19)`)
- Add `nemo_id` + `nemo_pfix` in class writer (`r pr(21)`, `r iss(18)`)
  - changed odir to diro arg in nemofy
- list_files: `str_remove` offers more flexibility (`r pr(24)`, `r iss(23)`)
- Add metadata export functionality (`r pr(25)`, `r iss(17)`)
  - rename `nemo_id` + `nemo_pfix` to `input_id` + `input_pfix`, respectively
  - add `output_id`
- Bugfix: use `I()` for vroom literal data in `empty_tbl` (`r iss(27)`, `r pr(28)`)

## v0.0.1 (2025-09-07)

Initial release of nemo (`r pr(5)`)

- R pkg skeleton
- Documentation via pkgdown
- CLI
- Logger
- GitHub Actions CI/CD

Fixing issues `r iss(2)`, `r iss(3)` and `r iss(4)`
