---
title: "NEWS"
vignette: >
  %\VignetteIndexEntry{News}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: pkgs
#| message: false
#| warning: false
#| echo: false
{
  use("htmltools", "a")
  use("glue", "glue")
}
```

```{r echo=FALSE}
repo <- "https://github.com/tidywf/nemo"
vdiff <- function(new, old) {
  stopifnot(utils::compareVersion(new, old) == 1)
  htmltools::a(
    href = glue::glue("{repo}/compare/v{old}...v{new}"),
    glue::glue("v{new} - v{old} diff")
  )
}

pr_or_issue <- function(n, type) {
  x <- c(iss = "issues", pr = "pull")
  stopifnot(is.numeric(n), n > 0, type %in% names(x))
  type2 <- x[type]
  htmltools::a(
    href = glue::glue("{repo}/{type2}/{n}"),
    glue::glue("{type}{n}")
  )
}

iss <- function(n) {
  pr_or_issue(n, "iss")
}

pr <- function(n) {
  pr_or_issue(n, "pr")
}
```

## dev (2026-04-26)

- GitHub Actions: refactor deploy workflow for dev and main branches (`r pr(34)`)
- Schema simplification: replace separate `raw.yaml` + `tidy.yaml` per tool with a
  single unified `schema.yaml`. Each table entry now has a flat `columns:` list where
  each column carries its raw name, tidy name, type, description, and `since` version.
  `Config` derives versioned raw and tidy views from this unified source on demand.
- `Tool`: refactor parse/tidy dispatch — subclasses can now override per-table
  `parse_<tname>()` and `tidy_<tname>()` methods; falls back to ftype-based dispatch
  otherwise. Add support for `txt-nohead` and `csv-nohead-long` file types.
- `Tool1`: add `table4` (txt-nohead) and `table5` (csv-nohead-long) as examples of
  the new file type support.
- `Workflow1`: add to CLI workflow dispatcher (`nemoverse_wf_dispatch`) for
  end-to-end CLI testing.
- `inst/scripts/file_to_yaml.R`: refactor to generate `schema.yaml` skeletons from
  sample files; output goes to `inst/config/tools/_tmp/` to avoid overwriting real schemas.

## v0.0.3 (2026-02-09)

- `nemo_write`: use NA outpath attr for db

## v0.0.2 (2026-02-04)

- Add key-value parser (`r pr(12)`, `r iss(11)`)
- Export config writer (`r pr(13)`)
- Add table description to schema getters, contribution docs, pkgdown fixes (`r pr(19)`)
- Add `nemo_id` + `nemo_pfix` in class writer (`r pr(21)`, `r iss(18)`)
  - changed odir to diro arg in nemofy
- list_files: `str_remove` offers more flexibility (`r pr(24)`, `r iss(23)`)
- Add metadata export functionality (`r pr(25)`, `r iss(17)`)
  - rename `nemo_id` + `nemo_pfix` to `input_id` + `input_pfix`, respectively
  - add `output_id`
- Bugfix: use `I()` for vroom literal data in `empty_tbl` (`r iss(27)`, `r pr(28)`)

## v0.0.1 (2025-09-07)

Initial release of nemo (`r pr(5)`)

- R pkg skeleton
- Documentation via pkgdown
- CLI
- Logger
- GitHub Actions CI/CD

Fixing issues `r iss(2)`, `r iss(3)` and `r iss(4)`
