---
title: "NEWS"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
    echo: false
---

```{r}
#| label: pkgs
#| message: false
#| warning: false
{
  use("htmltools", "a")
  use("glue", "glue")
}
```

```{r}
#| label: setup
repo <- "https://github.com/tidywf/nemo"
vdiff <- function(new, old) {
  stopifnot(utils::compareVersion(new, old) == 1)
  htmltools::a(
    href = glue::glue("{repo}/compare/v{old}...v{new}"),
    glue::glue("v{new} - v{old} diff")
  )
}

pr_or_issue <- function(n, type) {
  x <- c(iss = "issues", pr = "pull")
  stopifnot(is.numeric(n), n > 0, type %in% names(x))
  type2 <- x[type]
  htmltools::a(
    href = glue::glue("{repo}/{type2}/{n}"),
    glue::glue("{type}{n}")
  )
}

iss <- function(n) {
  pr_or_issue(n, "iss")
}

pr <- function(n) {
  pr_or_issue(n, "pr")
}
```

## dev

- Add [DVC](https://dvc.org/) support, mostly for child packages

## v0.1.0 (2026-06-22)

Major refactor.

### Breaking changes

**Unified schema format** (`r pr(68)`)

- `schema.yaml` replaces the separate `raw.yaml` + `tidy.yaml` pair per tool.
  Each column now carries `raw`, `tidy`, `type`, `description`, and `versions` in
  one place under a flat `tables:` map. Child packages must migrate their per-tool
  configs to this format.

**Config / Tool / Workflow API overhaul** (`r pr(64)`, `r pr(65)`, `r pr(66)`)

- `Config`: `get_raw_*` / `get_tidy_*` methods renamed (e.g. `get_raw_patterns()`
  → `get_patterns()`, `get_raw_schemas_all()` → `get_schemas_raw()`);
  `are_raw_schemas_valid()` removed (validation now in `initialize()`);
  `config`, `raw_schemas_all`, `tidy_schemas_all` fields moved to private;
  `pkg` added as a public field; `cloneable = FALSE`
- `Tool`: `$files`, `$tbls`, `$files_tbl` moved to private — use `list_files()` /
  `get_tbls()` instead; schema delegate fields (`get_raw_schema`, `get_tidy_schema`)
  removed — use `tool$config$get_schema_raw()` / `tool$config$get_schema_tidy()`;
  `nemofy(diro = ...)` → `run(output_dir = ...)`;
  column `group` → `prefix_suffix`; `cloneable = FALSE`
- `Workflow`: `$tools`, `$files_tbl` moved to private — use `get_tools()`;
  `nemofy(diro = ...)` → `run(output_dir = ...)`; `list_files(type = ...)` — `type`
  arg removed; gains `metapkg` constructor argument; path existence validated at
  construction; `filter_files()` now validates names against known `tool_parser` values;
  `cloneable = FALSE`

**CLI renames**

- `--out_dir` → `--output_dir`; `--pfix_include` → `--prefix_include`;
  output column `input_pfix` → `input_prefix`; `--prefix_include` is now opt-in
  (no longer added by default)

**Other**

- `nemo_metadata()`: parameter `input_dir` => `input_dirs`; return value changed
  from a named list to a single-row tibble with list-columns
- `RPostgres` dropped as a hard dependency - callers now supply `dbdrv` explicitly

### New features

- New assertion module: `nemo_stop()`, `nemo_assert_scalar_chr()`, `nemo_assert_chr()`,
  `nemo_assert_not_null()`, `nemo_assert_out_fmt()` replacing `assertthat` calls
- Schema scaffold helpers: `config_prep_raw_schema()`, `config_prep_raw()`,
  `config_prep_multi()`, `config_prep_write()` — bootstrap a `schema.yaml` from
  example raw files
- `nemo_gha_mermaid()`: generates a Mermaid flowchart of the full CI/CD pipeline (`r pr(52)`)
- `nemo_uml()`: generates a PlantUML SVG from R6 class names
- `nemo_schema_reactable()` / `nemo_schemavis_data()`: interactive schema explorer
- New file types: `txt-nohead`, `txt-keyvalue`, `csv`, `csv-nohead-long`
- `tidy`: `--output_id` / `--ulid` flags for tagging outputs with a run identifier;
  `--max` flag on `list` subcommand
- Metadata exported to `metadata.parquet` per run (`r pr(62)`, `r iss(61)`)
- `Tool` and `Workflow` now both support `write_metadata` boolean

### Performance

- File discovery ~20× faster on large directories — benchmarked at 1.9s => 0.1 s
  on 43k files x 15 patterns

### Infrastructure

- CI/CD: `deploy.yaml` refactored from a 136-line monolith to reusable workflows
  via `tidywf/actions`; now also triggers on `dev` branch pushes
- `@claude` GHA restricted to repo owners, collaborators, and members
- `dependabot.yml` added for automated GHA dependency updates
- conda: aarch64 lock file added; dedicated `bump.yaml` env added

## v0.0.3 (2026-02-09)

- `nemo_write`: use NA outpath attr for db

## v0.0.2 (2026-02-04)

- Add key-value parser (`r pr(12)`, `r iss(11)`)
- Export config writer (`r pr(13)`)
- Add table description to schema getters, contribution docs, pkgdown fixes (`r pr(19)`)
- Add `nemo_id` + `nemo_pfix` in class writer (`r pr(21)`, `r iss(18)`)
  - changed odir to diro arg in nemofy
- list_files: `str_remove` offers more flexibility (`r pr(24)`, `r iss(23)`)
- Add metadata export functionality (`r pr(25)`, `r iss(17)`)
  - rename `nemo_id` + `nemo_pfix` to `input_id` + `input_pfix`, respectively
  - add `output_id`
- Bugfix: use `I()` for vroom literal data in `empty_tbl` (`r iss(27)`, `r pr(28)`)

## v0.0.1 (2025-09-07)

Initial release of nemo (`r pr(5)`)

- R pkg skeleton
- Documentation via pkgdown
- CLI
- Logger
- GitHub Actions CI/CD

Fixing issues `r iss(2)`, `r iss(3)` and `r iss(4)`
