---
title: "Schema"
vignette: >
  %\VignetteIndexEntry{Schema}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
#| include: false
use("nemo")
```

Each tool has a single [LinkML](https://linkml.io/linkml/index.html) schema under
`inst/config/tools/<tool>/schema.yaml`. This file tells {nemo} everything it
needs to know: which files to look for, how to parse them, and what the tidy output
should look like.

## File header

Every schema starts with the same boilerplate:

```yaml
id: https://example.org/nemo/tool1
name: tool1
description: >-
  Just a toy tool for demoing.

prefixes:
  linkml: https://w3id.org/linkml/
  tool1: https://example.org/nemo/tool1/

imports:
  - linkml:types

default_range: string
default_prefix: tool1
```

- `id`, `name`, `description`, `prefixes`, and `imports`: LinkML conventions.
- `default_range: string`: use string as a fallback for any attribute without an
  explicit `range`.

## Subsets

`subsets` declares all the tags that will be referenced in `in_subset` later:

- `raw`: class describes **raw input** file.
- `tidy`: class describes **tidy output** tibble.
- `<version>`: any other name (e.g. `v1.2.3`, `latest`) tags a version.

```yaml
subsets:
  raw:
    description: Raw input file schema
  tidy:
    description: Tidy output schema
  v1.2.3:
    description: Tool1 version v1.2.3
  latest:
    description: Tool1 latest version
```

At least `raw`, `tidy`, and one version subset (typically `latest`) are
always required.

## Class naming convention

Classes under `classes:` must be named with a `Raw` or `Tidy` prefix.
The prefix is stripped and lowercased to get the **logical table name** used
in the Config API. For example, `RawTable1` and `TidyTable1` both refer to
the logical table `table1`.

To decouple the class name from the table name (useful when a single table
maps to multiple tidy subtables), set the `name` annotation explicitly, which
overrides the prefix-stripping fallback.

## Visualisation

Schemas can be visualised as Mermaid ER diagrams using `schema_to_mermaid()`.
All schema versions are merged into a single view.

For example, the `tool1` schema looks like:

```{r tool1-mermaid}
#| results: asis
#| echo: true
p <- system.file("config/tools/tool1/schema.yaml", package = "nemo")
cat("```{mermaid}\n", schema_to_mermaid(p), "\n```\n", sep = "")
```

## Annotations

Every class carries a set of {nemo}-specific `annotations`:

| Key       | Required on       | Purpose                                             |
|-----------|-------------------|-----------------------------------------------------|
| `name`    | both Raw and Tidy | Logical table name; overrides prefix-strip fallback |
| `pattern` | Raw only          | Regex used to find matching files in a directory    |
| `ftype`   | Raw only          | File type (`tsv`, `csv`, `txt-nohead`)              |
| `subtbl`  | Tidy only         | Sub-table name within a parser (default: `tbl1`)    |

## Field types

LinkML `range` maps to {nemo}'s internal readr col-spec code:

- string: character (`c`)
- integer: integer (`i`)
- float: double (`d`)

## Corner cases

### 1. Simple 1:1 mapping (single version)

The common case: one raw file, one tidy tibble, latest version only. Both
classes carry `in_subset: [raw, latest]` / `[tidy, latest]` at the class level,
which applies to all their attributes.

```yaml
subsets:
  raw:
    description: Raw input file schema
  tidy:
    description: Tidy output schema
  latest:
    description: Tool1 latest version

classes:
  RawTable2:
    description: Table2 raw file.
    in_subset: [raw, latest]
    annotations:
      name: table2
      pattern: "\\.tool1\\.table2\\.tsv$"
      ftype: tsv
    attributes:
      SampleID:
        range: string
      metricA:
        range: string
      metricB:
        range: float

  TidyTable2:
    description: Table2 tidy output.
    in_subset: [tidy, latest]
    annotations:
      name: table2
      subtbl: tbl1
    attributes:
      sample_id:
        range: string
        description: Sample ID
      metric_a:
        range: string
        description: Metric A
      metric_b:
        range: float
        description: Metric B
```

The `tool1` schema included with {nemo} demonstrates this pattern for `table2`.
We can inspect it programmatically:

```{r}
#| label: simple-schema
conf <- Config$new("tool1", pkg = "nemo")
conf$get_raw_schema("table2")
conf$get_tidy_schema("table2")
```

### 2. Multi-version schema (slot-level `in_subset`)

When a tool's output schema changes between versions, **slot-level**
`in_subset` tags declare which fields belong to which version. The class-level
`in_subset` should include **only** `raw` (or `tidy`); version membership is
inferred from the slots.

```yaml
classes:
  RawTable1:
    description: Table1 raw file.
    in_subset:
      - raw            # <-- version tags omitted; slots carry them
    annotations:
      name: table1
      pattern: "\\.tool1\\.table1\\.tsv$"
      ftype: tsv
    attributes:
      SampleID:
        range: string
        in_subset: [v1.2.3, latest]   # present in both versions
      Chromosome:
        range: string
        in_subset: [v1.2.3, latest]
      Start:
        range: integer
        in_subset: [v1.2.3, latest]
      End:
        range: integer
        in_subset: [v1.2.3, latest]
      metricX:
        range: float
        in_subset: [v1.2.3, latest]
      metricY:
        range: float
        in_subset: [latest]           # added after v1.2.3
      metricZ:
        range: float
        in_subset: [latest]           # added after v1.2.3
```

The Config API handles version resolution automatically:

```{r}
#| label: versioned-schema
conf$get_raw_schema("table1", v = "v1.2.3") # 5 fields
conf$get_raw_schema("table1", v = "latest") # 7 fields
conf$get_raw_versions()
```

The `schema_guess()` function (called internally by `parse_file()`) uses
exact column-count and column-name matching to select the correct version
for any given file automatically.

### 3. No-header file (`ftype: txt-nohead`)

Some tools write key-value or fixed-column files without a header row. Set
`ftype: txt-nohead` and list the column names in positional order. The parser
assigns names in the order they appear in `attributes`.

```yaml
  RawTable3:
    description: Table3 raw file (no header).
    in_subset: [raw, latest]
    annotations:
      name: table3
      pattern: "\\.tool1\\.table3\\.tsv$"
      ftype: txt-nohead
    attributes:
      SampleID:
        range: string
      QCStatus:
        range: string
      TotalReads:
        range: float
      MappedReads:
        range: float
      UnmappedReads:
        range: float
```

The corresponding `tool1/table3.tsv` file has no header, where columns are bound
by position:

```{r nohead-schema}
conf$get_raw_schema("table3")
```

### 4. Multiple tidy subtables from one parser

Sometimes a single raw file is parsed into multiple tidy tibbles (e.g. for
normalisation or because the file mixes different entity types). Use a
distinct `subtbl` annotation on each tidy class; both share the same `name`.

```yaml
  TidySummaryTbl1:
    description: Summary metrics.
    in_subset: [tidy, latest]
    annotations:
      name: summary       # same logical name
      subtbl: tbl1        # different subtable
    attributes:
      metric_a:
        range: float
        description: Metric A
      metric_b:
        range: float
        description: Metric B

  TidySummaryCovx:
    description: Coverage per depth bucket.
    in_subset: [tidy, latest]
    annotations:
      name: summary       # same logical name
      subtbl: tbl2        # different subtable
    attributes:
      covx:
        range: integer
        description: Coverage depth threshold
      value:
        range: float
        description: Fraction of bases >= covx
```

The tidy method retrieves each subtable by name:

```r
schema_tbl1 <- self$get_tidy_schema("summary", subtbl = "tbl1")
schema_tbl2 <- self$get_tidy_schema("summary", subtbl = "tbl2")
```

### 5. Custom parser with a dummy raw field

When the raw file format is too irregular for the generic `readr`-based
parser (e.g. `samtools flagstat` output), the Tool subclass implements its
own `parse_*` method. The raw class in the schema still needs to exist for
file discovery (pattern matching), but its attributes section is a single
placeholder:

```yaml
  RawFlagstats:
    description: Counts of alignments for each FLAG type.
    in_subset: [raw, latest]
    annotations:
      name: flagstats
      pattern: "\\.flagstat$"
      ftype: txt
    attributes:
      dummy:
        range: string
```

The tidy class is written normally, where the custom `parse_flagstats()` method
returns a tibble whose columns match `TidyFlagstats` exactly.

### 6. Field names with dots

YAML keys that contain a dot (`.`) must be quoted, otherwise YAML parses
them incorrectly:

```yaml
    attributes:
      "build.date":
        range: string
```

This applies to any special character in a field name. Without quotes the key
will be silently misread.

### 7. Intermediate tidy form (raw rename before custom split)

When a Tool's `tidy_*` method calls `self$.tidy_file(x, "parser")` as an
intermediate step before custom R code splits or reshapes the result, the
`Tidy*` class must declare **all columns present after the rename step**
— not just the final output columns.

For example, `tidy_genecvg()` calls `.tidy_file()` (which renames all
columns from CamelCase to snake_case) and then splits the wide tibble into
two narrow tibbles (genes and per-base coverage). The schema's `TidyGenecvg`
class must list all ~70 renamed columns (which sucks I know).

```yaml
  TidyGenecvg:
    # All ~70 post-rename columns must be listed here so that
    # get_tidy_schema() can build the correct readr column spec.
    in_subset: [tidy, latest]
    annotations:
      name: genecvg
      subtbl: tbl1
    attributes:
      gene:
        range: string
        description: Gene name
      chrom:
        range: string
        description: Chromosome
      pos_start:
        range: integer
        description: Start position
      # ...
```

- Rule of thumb: `Tidy*` class must match whatever `colnames()` the
tibble has at the point where `get_tidy_schema()` is called in the
`tidy_*` method.
