---
title: "Schema"
vignette: >
  %\VignetteIndexEntry{Schema}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
#| include: false
use("nemo")
use("dplyr", c("select", "filter", "mutate"))
use("knitr", "kable")
```

Each nemo tool has a single `schema.yaml` under `inst/config/tools/<tool>/`.
It is a flat `tables:` map — one entry per output file type — that drives both
file discovery and raw => tidy transformation.

## `schema.yaml` structure

```yaml
tables:
  <table_name>:
    description: '<human-readable description>'
    pattern: "<regex matching the file basename>"
    ftype: '<file type — see below>'
    columns:
      - raw: '<raw column name or positional label>'
        tidy: '<tidy snake_case name>'
        type: 'char | int | float'
        description: '<human-readable description>'
        since: '<version string, e.g. v1.2.3 or latest>'
```

The `pattern` value must use YAML **double-quoted** strings so that backslash
escapes are processed correctly (e.g. `"\\.tool1\\.table1\\.tsv$"`).

## File types (`ftype`)

The `ftype` field signals the parsing strategy for that table.
Tab-delimited is the default; other delimiters are handled inside the
specific parser, not in `ftype`.

| `ftype`           | Description                        | `raw:` names        | Parse function |
|-------------------|------------------------------------|---------------------|----------------|
| `txt`             | Header present; cols named in file | actual column names | `parse_file` |
| `txt-nohead`      | No header; positional columns      | `X1, X2, …, XN`     | `parse_file_nohead` |
| `txt-keyvalue`    | No header; exactly 2 cols (key, value); pivoted wide | actual key values | `parse_file_keyvalue` |
| `csv-nohead-long` | No header; long format with a metric-name col; custom pivot logic | metric names (values in the key column) | custom `tidy_*` method |

### `txt`

Standard delimited file with a header row.
The `raw:` names must exactly match the column names in the file — they are used
by `schema_guess` to identify which versioned schema applies to a given file.

### `txt-nohead`

No header row; columns are identified purely by position.
The `raw:` names are always `X1, X2, …, XN` (where N = number of columns).
This makes version tracking unambiguous: adding or removing a column is
reflected by a new `since:` entry for the affected positional label.
The tidy step renames `X1 => tidy_name_1`, `X2 => tidy_name_2`, etc. positionally.

### `txt-keyvalue`

A special case of `txt-nohead` with exactly two columns where the first column
is a key and the second is its value.
The file is pivoted wide so that each key becomes a column name.
The `raw:` names are the actual key values (e.g. `SampleID`, `TotalReads`),
which are matched by `schema_guess` after the pivot.

### `csv-nohead-long`

No header; long format where one column holds metric names and adjacent columns
hold their values (e.g. count and percentage).
The `columns:` block only maps raw metric names to tidy column names.
The structural logic (which column is the key, `_pct` suffix generation, section
cleaning) lives in the tool's `tidy_*` method, not in the schema.

## Versioning (`since:`)

The `since:` field on each column records the **first version** of the tool that
produced that column.
The effective schema for a file at version `V` is all columns where `since <= V`,
with `latest` always treated as the highest version.

```yaml
table1 columns:
  SampleID   since: v1.2.3   # present in v1.2.3 and later
  Chromosome since: v1.2.3
  Start      since: v1.2.3
  End        since: v1.2.3
  metricX    since: v1.2.3
  metricY    since: latest   # added in latest
  metricZ    since: latest
```

`schema_guess` uses this to automatically match a file's column names against the
correct versioned schema.
Tables with a single version (all columns `since: latest`) are the common case
and require no special handling.

## Live example

`Config$new()` reads the `schema.yaml` and exposes the derived schemas:

```{r}
conf <- Config$new("tool1", pkg = "nemo")
conf
```

### Patterns and file types

```{r}
conf$get_patterns() |> kable()
```

```{r}
conf$get_ftypes() |> kable()
```

### Versioned raw schemas (used by `schema_guess`)

```{r}
conf$get_schemas_all() |>
  select("name", "version", "schema")
```

The `schema` list-col holds the `(field, type)` pairs used when reading a file:

```{r}
conf$get_schema("table1", v = "v1.2.3", raw_or_tidy = "raw") |>
  select("field", "type") |>
  kable()
```

### Tidy schema (used for column renaming)

```{r}
conf$get_schema("table1", raw_or_tidy = "tidy") |> kable()
```

### Column map for `csv-nohead-long` tables

For `csv-nohead-long` tables, `get_col_map()` returns the metric-name mapping
used inside the `tidy_*` method:

```{r}
conf$get_col_map("table5") |> kable()
```
