---
title: "Schema Walkthrough"
vignette: >
  %\VignetteIndexEntry{Schema Walkthrough}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: pkgs
#| include: false
use("nemo", c("Config"))
use("dplyr", c("select", "filter", "mutate"))
use("knitr", "kable")
```

Each tool has a single `schema.yaml` under `inst/config/tools/<tool>/`.
It is a flat `tables` map, with one entry per output file that drives both
file discovery and raw-to-tidy transformation.

```yaml
tables:
  <table_name>:
    description: '<human-readable description>'
    pattern: "<regex matching the file basename>"
    ftype: '<file type — see below>'
    columns:
      - raw: '<raw column name or positional label>'
        tidy: '<tidy snake_case name>'
        type: 'char | int | float'
        description: '<human-readable description>'
        versions: ['<version1>', '<version2>']
```

## Table name

We keep table names as all lowercase, alphanumeric characters. Example: `table1`.

## Description (`description`)

A short description for the table that can be used to summarise the contents
of the raw file. Example: `Table foo contains one row per somatic variant.`.

## Pattern (`pattern`)

A character string containing a regular expression to be matched against
each discovered file's basename. You must use **double-quoted** strings in
the YAML so that backslash escapes are processed correctly.
Example: `"\\.tool1\\.table1\\.tsv$"`.

## File type (`ftype`)

The `ftype` field signals the parsing strategy for that table, including the
delimiter. Tab-delimited files use `txt` or `txt-nohead`; comma-delimited files
with a header use `csv`. More exotic formats require a custom `parse_*` method.

| `ftype`           | Description                        | `raw:` names        | Parse function |
|-------------------|------------------------------------|---------------------|----------------|
| `txt`             | Header present; tab-delimited      | actual column names | `parse_file` |
| `csv`             | Header present; comma-delimited    | actual column names | `parse_file` (delim=`,`) |
| `txt-nohead`      | No header; positional columns      | `X1, X2, ..., XN`   | `parse_file_nohead` |
| `txt-keyvalue`    | No header; exactly 2 cols (key, value); pivoted wide | actual key values | `parse_file_keyvalue` |
| `csv-nohead-long` | No header; long format with a metric-name col; custom pivot logic | metric names (values in the key column) | custom `tidy_*` method |

### `txt`

Standard tab-delimited file with a header row.
The `raw:` names must exactly match the column names in the file — they are used
by `nemo::schema_guess` to identify which versioned schema applies to a given file.

### `csv`

Same as `txt` but comma-delimited. Auto-dispatched to `parse_file` with
`delim = ","`. Use this for any tool that produces `.csv` output files with a
header row.

### `txt-nohead`

No header row; columns are identified purely by position.
The `raw` names are always `X1, X2, ..., XN` (where N = number of columns).
This makes version tracking unambiguous: adding or removing a column is
reflected by updating the `versions` array for the affected positional label.
The tidy step renames `X1 => tidy_name_1`, `X2 => tidy_name_2`, etc. positionally.

### `txt-keyvalue`

A special case of `txt-nohead` with exactly two columns where the first column
is a key and the second is its value.
The file is pivoted wide so that each key becomes a column name.
The `raw` names are the actual key values (e.g. `SampleID`, `TotalReads`),
which are matched by `nemo::schema_guess` after the pivot.

### `csv-nohead-long`

No header; long format where one column holds metric names and adjacent columns
hold their values (e.g. count and percentage).
The `columns` block only maps raw metric names to tidy column names.
The structural logic (which column is the key, `_pct` suffix generation, section
cleaning) lives in the tool's `tidy_*` method, not in the schema.

## Versioning (`versions`)

The `versions` field for each column is an explicit array of every tool version
that column appears in. This gives a full picture of column history, including
additions and removals across versions.

```yaml
table1 columns:
  SampleID   versions: ['v1.2.3', 'v4.5.6', 'latest']  # present in all versions
  Chromosome versions: ['v1.2.3', 'v4.5.6', 'latest']
  Start      versions: ['v1.2.3', 'v4.5.6', 'latest']
  End        versions: ['v1.2.3', 'v4.5.6', 'latest']
  metricX    versions: ['v1.2.3']                      # dropped in v4.5.6
  metricY    versions: ['latest']                      # added in latest
  metricZ    versions: ['latest']
```

The effective schema for version `V` is all columns whose `versions` array
contains `V`. `schema_guess` uses this to automatically match a file's column
names against the correct versioned snapshot.

## Examples

`nemo::Config$new()` reads the `schema.yaml` and exposes the derived schemas:

```{r}
#| label: config_new
conf <- Config$new("tool1", pkg = "nemo")
conf
```

### Patterns and file types

```{r}
#| label: get_pats
conf$get_patterns() |> kable()
```

```{r}
#| label: get_ftypes
conf$get_ftypes() |> kable()
```

### Versioned raw schemas (used by `nemo::schema_guess`)

```{r}
#| label: get_schemas_all
conf$get_schemas_all() |>
  select("name", "version", "schema")
```

The `schema` list-col holds the `(field, type)` pairs used when reading a file:

```{r}
#| label: get_schema_table1_raw
conf$get_schema("table1", v = "v1.2.3", raw_or_tidy = "raw") |>
  select("field", "type") |>
  kable()
```

### Tidy schema (used for column renaming)

```{r}
#| label: get_schema_table1_tidy
conf$get_schema("table1", raw_or_tidy = "tidy", v = "latest") |> kable()
```

### Column map for `csv-nohead-long` tables

For `csv-nohead-long` tables, `get_col_map()` returns the metric-name mapping
used inside the `tidy_*` method:

```{r}
#| label: get_col_map_table5
conf$get_col_map("table5") |> kable()
```
