---
title: "Structure"
output: rmarkdown::html_document
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
params:
  tool: "Tool1"
  workflow: "Workflow1"
  pkg: "nemo"
---

```{css, css-chunk, echo=FALSE}
/*
https://github.com/r-lib/pkgdown/issues/2721
Correct flex boxes from sizing based upon the largest in tabset
*/

.tab-content>.tab-pane {
  display: none;
}

.tab-content>.active {
  display: block;
}
```

```{r}
#| label: params_eval
#| echo: false
#| eval: false

params <- list(
  tool = "Tool1",
  workflow = "Workflow1",
  pkg = "nemo"
)
```

```{r}
#| label: pkgs
#| message: false
#| warning: false
#| echo: false

{
  use(params$pkg)
  use("dplyr")
  use("glue", "glue")
  use("purrr", "map")
  use("readr", "read_tsv")
  use("tibble", "tibble")
  use("tidyr", "unnest")
  use("fs", "dir_info")
  use("knitr", "kable")
  use("arrow", "read_parquet")
}
```

```{r}
#| label: opts
#| echo: false

options(width = 100)
```

nemo has three base [R6] classes: `Config`, `Tool`, `Workflow`. Child packages
([tidywigits], [tidydragen]) subclass `Tool`/`Workflow` with tool-specific
parsers and tidiers.

[R6]: <https://github.com/r-lib/R6> "R6 package repo"
[tidywigits]: <https://github.com/tidywf/tidywigits> "tidywigits"
[tidydragen]: <https://github.com/tidywf/tidydragen> "tidydragen"

Examples below use the built-in `Tool1` and `Workflow1`.

## Config {.tabset .tabset-pills}

Reads `inst/config/tools/<tool>/schema.yaml` and serves versioned raw/tidy
schemas from it. See `?Config`.

### format

One `tables` entry per output file:

```yaml
tables:
  <table_name>:
    description: '<human-readable description>'
    pattern: "<regex matching the file basename>"
    glob: ["<aws s3 sync --include glob>", '...']
    ftype: '<file type — see below>'
    columns:
      - raw: '<raw column name or positional label>'
        tidy: '<tidy snake_case name>'
        type: 'char | int | float'
        description: '<human-readable description>'
        versions: ['<version1>', '<version2>']
```

- `pattern`: PCRE regex on the file **basename**; `Tool` uses it to discover and
  classify local files.
- `glob`: list of `aws s3 sync --include` patterns on the whole key;
  `wf_sync_patterns()` collects them to decide what to sync down. Usually the
  basename pattern with a leading `*` (`"\\.tool1\\.table1\\.tsv$"` =>
  `["*.tool1.table1.tsv"]`).

`ftype` picks the parser:

| `ftype`        | Description                                  | `raw:` names        | Parser                   |
| -------------- | -------------------------------------------- | ------------------- | ------------------------ |
| `tsv`          | Header present; tab-delimited                | actual column names | `parse_file`             |
| `csv`          | Header present; comma-delimited              | actual column names | `parse_file` (delim=`,`) |
| `tsv-nohead`   | No header; positional columns                | `X1, X2, …, XN`     | `parse_file_nohead`      |
| `tsv-keyvalue` | No header; 2 cols (key, value); pivoted wide | actual key values   | `parse_file_keyvalue`    |

Child packages add ftypes by overriding `private$extra_ftypes()` (named list of
`ftype -> function(x, table_name)`).

`versions` lists every tool version a column appears in:

```yaml
# table1 columns
SampleID:   versions: ['v1.2.3', 'v4.5.6', 'latest']  # present in all versions
metricX:    versions: ['v1.2.3']                        # dropped in v4.5.6
metricY:    versions: ['latest']                        # added in latest
```

Schema for version `V` = columns whose `versions` contain `V`. `schema_guess`
matches a file's columns to the right version.

### intro

```{r}
#| label: config_intro

tool <- params$tool
workflow <- params$workflow
conf <- nemo::Config$new(tool, pkg = "nemo")
conf
```

### tables

All tables:

```{r}
#| label: config_tables_all

conf$get_patterns() |> knitr::kable(caption = glue("{tool} patterns."))
conf$get_globs() |> knitr::kable(caption = glue("{tool} globs."))
conf$get_ftypes() |> knitr::kable(caption = glue("{tool} file types."))
conf$get_descriptions() |> knitr::kable(caption = glue("{tool} descriptions."))
```

One table:

```{r}
#| label: config_tables_one

conf$get_pattern("table1")
conf$get_ftype("table1")
conf$get_description("table1")
```

### schemas

All versions:

```{r}
#| label: config_schemas_all

conf$get_schemas_raw() |> dplyr::select("name", "version", "schema")
conf$get_schemas_tidy() |> dplyr::select("name", "version", "schema")
conf$get_schemas_both()
```

One table, one version:

```{r}
#| label: config_schemas_one

conf$get_schema_raw("table1", version = "v1.2.3")
conf$get_schema_tidy("table1", version = "v1.2.3")
```

Raw-to-tidy column map, for use in `tidy_*` methods:

```{r}
#| label: config_col_map

conf$get_col_map("table6")
```

## Tool {.tabset .tabset-pills}

Finds, parses, tidies and writes one tool's files. See `?Tool`.

### initialise

```{r}
#| label: tool_tool1

tool1_path <- system.file("extdata/tool1", package = "nemo")
t1 <- nemo::Tool1$new(path = tool1_path)
# each class comes with a print function
t1
```

### config

```{r}
#| label: tool1_conf

t1$config
t1$config$get_patterns()
t1$config$get_schema_raw("table1", version = "v1.2.3")
# t1$config$get_schema_raw("table1", version = "latest") # default
t1$config$get_schema_tidy("table1", version = "v1.2.3")
# t1$config$get_schema_tidy("table1", version = "latest") # default
```

### list

```{r}
#| label: tool1_list

lf <- t1$list_files()
lf |> dplyr::slice(1) |> str()
```

### tidy

`tidy()` modifies the object in place:

```{r}
#| label: tool1_tidy

# this will create a new field tbls containing the tidy data (and optionally
# the 'raw' parsed data)
t1$tidy(keep_raw = TRUE)
tbls <- t1$get_tbls()
tbls
tbls$raw[[1]] |> dplyr::glimpse()
# the tidy tibbles are nested to allow for more than one tidy tibble per file
tbls$tidy[[1]][["data"]][[1]] |> dplyr::glimpse()
```

### filter

Keep/drop `tool_parser`s with `include`/`exclude`:

```{r}
#| label: tool1_filter

# create new Tool1 object
t2 <- nemo::Tool1$new(path = tool1_path)
t2$list_files()
t2$filter_files(include = c("tool1_table2", "tool1_table3"))
t2$list_files()
```

### write

Formats: `parquet`, `db`, `tsv`, `csv`, `rds`. Optional `input_id`, `output_id`
and `input_prefix` (`prefix_include = TRUE`) columns get prepended. See [output
naming](output_naming.html).

```{r}
#| label: tool1_write

t2$tidy() # first need to tidy
outdir1 <- file.path(tempdir(), "tool1_write")
fmt <- "csv"
t2$write(
  output_dir = outdir1,
  format = fmt,
  input_id = "run123",
  output_id = "out123",
  prefix_include = TRUE
)
wfiles <- fs::dir_info(outdir1) |>
  dplyr::mutate(bname = basename(.data$path)) |>
  dplyr::select("bname", "size", "type")
wfiles
```

### run

`run()` = filter + tidy + write:

```{r}
#| label: run1

t3 <- nemo::Tool1$new(path = tool1_path)
outdir2 <- file.path(tempdir(), "t3")
t3$list_files()
t3$run(
  output_dir = outdir2,
  format = "tsv",
  input_id = "run_t3",
  output_id = "out456",
  prefix_include = TRUE
)
wfiles2 <- fs::dir_info(outdir2) |>
  dplyr::mutate(bname = basename(.data$path)) |>
  dplyr::select("path", "bname", "size", "type")
wfiles2 |> dplyr::select(-"path")
tbl1_path <- wfiles2 |>
  dplyr::filter(stringr::str_detect(.data$bname, "tool1_table1")) |>
  dplyr::pull("path") |>
  head(1)
readr::read_tsv(tbl1_path, show_col_types = FALSE)
```

## Workflow

A list of `Tool`s with the same `list_files`/`filter_files`/`tidy`/`write`/`run`
API, plus `get_tools()`. `Workflow1` wraps a single `Tool1`. See `?Workflow`.

```{r}
#| label: w1_init

w <- system.file("extdata/tool1", package = "nemo") |>
  nemo::Workflow1$new()
outdir3 <- file.path(tempdir(), "w1")
w$get_tools()
w$list_files()
x <- w$run(
  output_dir = outdir3,
  format = "tsv",
  input_id = "wf1_run1",
  output_id = "out1",
  prefix_include = TRUE
)
wfiles3 <- fs::dir_info(outdir3) |>
  dplyr::select(1:5) |>
  dplyr::mutate(bname = basename(.data$path))
wfiles3 |>
  dplyr::select("bname", "size", "type")
```

## Metadata

`Workflow$write()`/`$run()` (non-`db` formats) also write a single-row
`metadata.parquet`: IDs, paths, pkg versions, files written.

```{r}
#| label: metadata_read

meta <- arrow::read_parquet(file.path(outdir3, "metadata.parquet"))
meta |> dplyr::glimpse()
```

Unnest list columns:

```{r}
#| label: metadata_pkgs

meta |> dplyr::select("pkg_versions") |> tidyr::unnest("pkg_versions")
```

```{r}
#| label: metadata_files

meta |> dplyr::select("files") |> tidyr::unnest("files")
```
