---
title: "Structure"
output: rmarkdown::html_document
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
params:
  tool: "Tool1"
  workflow: "Workflow1"
  pkg: "nemo"
---

```{css, echo=FALSE}
/*
https://github.com/r-lib/pkgdown/issues/2721
Correct flex boxes from sizing based upon the largest in tabset
*/

.tab-content>.tab-pane {
  display: none;
}

.tab-content>.active {
  display: block;
}
```

```{r params_eval, echo=FALSE, eval=FALSE}
params <- list(
  tool = "Tool1",
  workflow = "Workflow1",
  pkg = "nemo"
)
```

```{r pkgs, message=FALSE, warning=FALSE, echo=FALSE}
{
  use(params$pkg)
  use("dplyr")
  use("glue", "glue")
  use("purrr", "map")
  use("readr", "read_tsv")
  use("tibble", "tibble")
  use("tidyr", "unnest")
  use("fs", "dir_info")
  use("knitr", "kable")
  use("arrow", "read_parquet")
}
```


```{r}
#| label: opts
#| echo: false
options(width = 100)
```

{nemo} is built on top of R's [R6] encapsulated object-oriented programming
implementation, which helps with code organisation. It consists of several base
classes like `Config`, `Tool`, and `Workflow` which we describe below.
Each R6 class can contain public and private functions and non-functions (fields).

Other R packages like {[tidywigits]} and {[tidydragen]} (in _dev_) can create their
own `Tool` and `Workflow` children classes that inherit (or override) functions from
the {nemo} parent classes. This allows for custom parsers and tidiers for specific
bioinformatic tools and workflows.

[R6]: <https://github.com/r-lib/R6> "R6 package repo"
[tidywigits]: <https://github.com/tidywf/tidywigits> "tidywigits"
[tidydragen]: <https://github.com/tidywf/tidydragen> "tidydragen"

Here we use the `Tool1` and `Workflow1` classes as examples to illustrate the
structure of the package.

## `Config` {.tabset .tabset-pills}

A `Config` object reads a single `schema.yaml` file from `inst/config/tools/<tool>/`.
Each table entry defines its file pattern, file type, description, and a list of
columns — each column carrying both its raw name (as produced by the tool) and its
tidy name (snake_case), type, description, and the `versions` array listing every
tool version the column appears in.
`Config` derives versioned raw and tidy views from this unified source on demand.
See `?Config`.

### intro

```{r tool1_config1}
tool <- params$tool
workflow <- params$workflow
conf <- nemo::Config$new(tool, pkg = "nemo")
conf
```

### patterns

Patterns are used to fish out the relevant files from a directory listing.

```{r tool1_config_patterns}
conf$get_patterns() |>
  knitr::kable(caption = glue("{tool} file patterns."))
```

### descriptions

Descriptions are usually derived from the individual tool documentation.

```{r tool1_config_descriptions}
conf$get_descriptions() |>
  knitr::kable(caption = glue("{tool} file descriptions."))
```

### versions

The `versions` array on each column lists every tool version that column appears
in, giving a full picture of additions and removals across versions.

```{r tool1_config_versions}
conf$get_schemas_raw() |>
  dplyr::select("name", "version") |>
  dplyr::distinct() |>
  knitr::kable(caption = glue("{tool} schema versions."))
```

### schemas

`get_schemas_raw()` and `get_schemas_tidy()` return versioned views with column
names in raw and tidy form respectively:

```{r tool1_config_schemas}
(s_raw <- conf$get_schemas_raw())
s_raw |>
  dplyr::filter(.data$name == "table1", .data$version == "v1.2.3") |>
  dplyr::select("schema") |>
  tidyr::unnest("schema")

(s_tidy <- conf$get_schemas_tidy())
s_tidy |>
  dplyr::filter(.data$name == "table1", .data$version == "v1.2.3") |>
  dplyr::select("schema") |>
  tidyr::unnest("schema")
```

## `Tool` {.tabset .tabset-pills}

`Tool` is the main organisation class for all file parsers and tidiers.
It contains functions for parsing and tidying typical CSV/TSV files (with
column names), and TXT files where the column names are missing. Currently it
utilises the very simple `readr::read_delim` function from the {[readr]} package
that reads all the data into memory. See `?Tool`.

These simple parsers are used in 80-90% of cases, so in the future we can
optimise the parsing if needed with faster packages such as {[data.table]},
{[duckdb-r]}/{[duckplyr]} or {[r-polars]}.

[readr]: <https://github.com/tidyverse/readr> "readr"
[duckdb-r]: <https://github.com/duckdb/duckdb-r> "duckdb-r"
[duckplyr]: <https://github.com/tidyverse/duckplyr> "duckplyr"
[data.table]: <https://github.com/Rdatatable/data.table> "data.table"
[r-polars]: <https://github.com/pola-rs/r-polars> "r-polars"

### initialise

We can have different `Tool` children classes that inherit (or override) functions
and fields from the `Tool` parent class.
For example, we can create a `Tool` object for `Tool1` as follows:

- Initialise a `Tool1` object:

```{r tool_tool1}
tool1_path <- system.file("extdata/tool1", package = "nemo")
t1 <- nemo::Tool1$new(path = tool1_path)
# each class comes with a print function
t1
```

### config

- Its `Config` object is also constructed based on the `name` supplied - this
is used internally to find files of interest and infer their schemas:

```{r tool1_conf}
t1$config
t1$config$get_patterns()
t1$get_schema_raw("table1", version = "v1.2.3")
# t1$get_schema_raw("table1", version = "latest") # default
t1$get_schema_tidy("table1", version = "v1.2.3")
# t1$get_schema_tidy("table1", version = "latest") # default
```

### list

We can list files that can be parsed with `list_files()`:

```{r tool1_list}
(lf <- t1$list_files())
lf |> dplyr::slice(1) |> str()
```

### tidy

We can parse and tidy files of interest using the `tidy` function. Note
that this function is called on the object and not assigned anywhere:

```{r tool1_tidy}
# this will create a new field tbls containing the tidy data (and optionally
# the 'raw' parsed data)
t1$tidy(do_tidy = TRUE, keep_raw = TRUE)
t1$tbls
t1$tbls$raw[[1]] |> dplyr::glimpse()
# the tidy tibbles are nested to allow for more than one tidy tibble per file
t1$tbls$tidy[[1]][["data"]][[1]] |> dplyr::glimpse()
```

### filter

We can also focus on a subset of files to tidy using the `filter_files()` function.
The `include` and `exclude` arguments can specify which `tool_parser`s to include or
exclude in the analysis:

```{r tool1_filter}
# create new Tool1 object
t2 <- nemo::Tool1$new(path = tool1_path)
t2$files
t2$filter_files(include = c("tool1_table2", "tool1_table3"))
t2$files
```

### write

After tidying the data of interest, we can write the tidy tibbles to various
formats, like Apache Parquet, PostgreSQL, CSV/TSV and R's RDS.
Below we can see that the (optional) `input_id` specified is added to the written
files in an additional `input_id` column. This can be used e.g. to distinguish results
from different samples in a data pipeline. An optional `output_id` column can also
be added (e.g. to distinguish between runs for the same input sample). An
`input_prefix` column can be included via `prefix_include = TRUE`, which would
use the prefix of the file basename.

```{r tool1_write}
t2$tidy() # first need to tidy
outdir1 <- tempdir()
fmt <- "csv"
t2$write(
  output_dir = outdir1,
  format = fmt,
  input_id = "run123",
  output_id = "out123",
  prefix_include = TRUE
)
wfiles <- fs::dir_info(outdir1) |> dplyr::select(1:5)
wfiles |>
  dplyr::mutate(bname = basename(.data$path)) |>
  dplyr::select("bname", "size", "type")
# readr::read_csv(wfiles$path[1], show_col_types = FALSE) # see tidywf/tidywigits#137
```

### wrangle

The `wrangle` function is a convenient wrapper for the process of filtering,
tidying, and writing.

```{r wrangle1}
t3 <- nemo::Tool1$new(path = tool1_path)
outdir2 <- file.path(tempdir(), "t3") |> fs::dir_create()
t3$files
t3$wrangle(
  output_dir = outdir2,
  format = "tsv",
  input_id = "run_t3"
)
wfiles2 <- fs::dir_info(outdir2) |>
  dplyr::mutate(bname = basename(.data$path))
wfiles2 |>
  dplyr::select("bname", "size", "type")
# no output_id or input_prefix specified
tbl1_path <- wfiles2$path[grep("tool1_table1", wfiles2$bname)[1]]
readr::read_tsv(tbl1_path, show_col_types = FALSE)
```

## `Workflow`

A `Workflow` consists of a list of one or more `Tool`s. We can construct a
certain `Workflow` with different `Tool`s, which would allow parsing and writing
tidy tables from a variety of bioinformatic tools. See `?Workflow`.

For example, {nemo} contains a `Workflow1` class as a `Workflow` child (containing
only a single `Tool1` for simplicity). Similarly to `Tool`, a `Workflow` object
contains functions such as `filter_files`, `list_files`, `tidy`, `write` and
`wrangle`:

```{r w1_init}
w <- system.file("extdata/tool1", package = "nemo") |>
  nemo::Workflow1$new()
outdir3 <- file.path(tempdir(), "oa") |> fs::dir_create()
w$list_files()
x <- w$wrangle(
  output_dir = outdir3,
  format = "tsv",
  input_id = "workflow1_run1",
  output_id = "out1",
  prefix_include = TRUE
)
wfiles3 <- fs::dir_info(outdir3) |>
  dplyr::select(1:5) |>
  dplyr::mutate(bname = basename(.data$path))
wfiles3 |>
  dplyr::select("bname", "size", "type")
```

## Metadata

After each `Workflow$write()` or `Workflow$wrangle()` call (for non-`db` formats),
a `metadata.parquet` file is written directly to the output directory alongside
the tidy tables.
It is a single-row tibble capturing the run context: IDs, input/output paths,
package versions, and the list of files written.

```{r metadata_read}
meta <- arrow::read_parquet(file.path(outdir3, "metadata.parquet"))
meta |> dplyr::glimpse()
```

The nested list columns can be unpacked with `tidyr::unnest()`:

```{r metadata_pkgs}
meta |> dplyr::select("pkg_versions") |> tidyr::unnest("pkg_versions")
```

```{r metadata_files}
meta |> dplyr::select("files") |> tidyr::unnest("files")
```
